---
title: "Linear Regression Models: Wine Quality Project"
date: "2022-12-15"
output:
  pdf_document: default
  html_document: default
group members: blank
editor_options: 
  markdown: 
    wrap: 72
---

#### **STAT GU4205/5205**

#### **Group Name: WQG**

##### **Group members: Gabriel Rios; gr2693, Triet Vo; tmv2123, Mingcong Cai; mc5258, Bihan Qian; bq2150, Amanda Samuel; acs2352**

#### Introduction

The Wine Quality project contains two datasets related to white and red
vinho verde wine samples, from the North of Portugal. The price of wine
depends on a rather abstract concept of wine appreciation by wine
tasters, opinions among whom may have a high degree of variability.
Physicochemical tests are key factors in wine certification and quality
assessment which are laboratory-based and take into account factors like
acidity, pH level, the presence of sugar, and other chemical properties.
For the wine market, it would be of interest if the human quality of
tasting can be related to the chemical properties of wine so that the
certification and quality assessment and assurance process is more
controlled.

Two datasets are available of which one dataset is on red wine and has
1599 different varieties and the other is on white wine and has 4898
varieties. All wines are produced in a particular area of Portugal. Data
are collected on 12 different properties of the wines one of which is
Quality, based on sensory data, and the rest are on chemical properties
of the wines including density, acidity, alcohol content etc. All
chemical properties of wines are continuous variables. Quality is an
ordinal variable with a possible ranking from 1 (worst) to 10 (best).
Each variety of wine is tasted by three independent tasters and the
final rank assigned is the median rank given by the tasters. Below are
columns that contained in our datasets:

Alcohol: the amount of alcohol in wine Volatile acidity: are high acetic
acid in wine which leads to an unpleasant vinegar taste Sulfates: a wine
additive that contributes to SO2 levels and acts as an antimicrobial and
antioxidant Citric Acid: acts as a preservative to increase acidity
(small quantities add freshness and flavor to wines) Total Sulfur
Dioxide: is the amount of free + bound forms of SO2 Density: sweeter
wines have a higher density Chlorides: the amount of salt in the wine
Fixed acidity: are non-volatile acids that do not evaporate readily pH:
the level of acidity Free Sulfur Dioxide: it prevents microbial growth
and the oxidation of wine Residual sugar: is the amount of sugar
remaining after fermentation stops. The key is to have a perfect balance
between --- sweetness and sourness (wines \> 45g/ltrs are sweet)

\newpage
#### Our main Research Question:

**Whether quality scores (impression scores from the human tasting) are
related to chemical compositions. Moreover, what physicochemical
properties influence the quality score of wine? Utilizing two datasets,
how would this relationship differ between red and white wine?**

In taking quality as the response variable, we are trying to understand
the variability in the wine quality through the variability of our
predictor variables. Moreover, our interest is focused on which
variables have the most impact on our response variable, wine quality,
as well as seeing if any one of our independent variables has an
influence over one another. Further, we want to know what measure of
variables would result in the best wine, and what constitutes the best
wine. We will be investigating this problem for both our red wine and
white wine datasets which will then allow us to understand the
physicochemical differences between red and white wine in relation to
their respective quality. By investigating and comparing our best
regression models for the two datasets, we will be able to accomplish
the above goal. 

Considering from more practical levels, our research based on the wine
quality dataset can contribute to the wine producers by lower pricing
cost and consumers in easier decision making. The idea of consumer
perspective stems from the 'paradox of choice' (Schwartz 2004) which
explores the psychology of decision-making. It introduces the idea that
an increase in consumer choices creates higher levels of anxiety for the
shopper. Schwartz presents the idea that one thing shoppers can do to
reduce their shopping anxiety is to have a clear understanding of what
they want. Unfortunately, developing an understanding of something as
intricate as wine quality takes time, money, and effort many consumers
cannot afford. For reference, the Institute of Culinary Education's
sommelier training program costs \~\$10,000 and consists of 200 hours of
instructional training. By applying regression analysis to the
quantifiable properties of wine quality we hope to lower this barrier to
entry and reduce shopper anxiety.

```{r setup, include=FALSE}
library(rdataretriever)
library(ggplot2)
library(rmarkdown)
library(car)
library(dplyr)
wine_quality_red_data <- read.csv(file="wine_quality_red.csv",as.is=T, header=T)
attach(wine_quality_red_data)

n <- data.frame(fixed_acidity = wine_quality_red_data$fixed_acidity, volatile_acidity = wine_quality_red_data$volatile_acidity, citric_acid = wine_quality_red_data$citric_acid, residual_sugar = wine_quality_red_data$residual_sugar, chlorides = wine_quality_red_data$chlorides, free_sulfur_dioxide = wine_quality_red_data$free_sulfur_dioxide, total_sulfur_dioxide = wine_quality_red_data$total_sulfur_dioxide, density = wine_quality_red_data$density, ph = wine_quality_red_data$ph, sulphates = wine_quality_red_data$sulphates, alcohol = wine_quality_red_data$alcohol, quality = wine_quality_red_data$quality)
```
\newpage
#### Variable transformations

We first use scatterplot Matrix for Red Wine and checking for
correlation between variables.
```{r, fig.width=15, fig.height=15}
plot(n)
```

Multicollinearity occurs when independent variables in a regression
model are correlated. This correlation is a problem because independent
variables should be independent. If the degree of correlation between
variables is high enough, it will lead to the inaccuracy of the whole
model. We need to be aware of it and dropping highly correlated
variables in our model building.

\newpage
Checking for Normality: Histogram & Q-Q Plot

```{r, echo = FALSE}
par(mar = c(2, 2, .5, .5))
par(mfrow = c(3,4))
hist(fixed_acidity, cex.main =.8)
hist(volatile_acidity, cex.main =.8)
hist(citric_acid, cex.main =.8)
hist(residual_sugar, cex.main =.8)
hist(chlorides, cex.main =.8)
hist(free_sulfur_dioxide, cex.main =.8)
hist(total_sulfur_dioxide, cex.main =.8)
hist(density, cex.main =.8)
hist(ph, cex.main =.8)
hist(sulphates, cex.main =.8)
hist(alcohol, cex.main =.8)
hist(quality, cex.main =.8)
```

Most 'normal' looking histograms: fixed_acidity, density, ph, and maybe
volatile acidity\
Right-skewed distributions: Volatile acidity, citric acid, residual
sugar, chlorides, free sulfur dioxide, total sulfur dioxide, sulphates,
alcohol.

Transforming predictors/Combine Predictors:

Log transformation improved: fixed_acidity, volatile_acidity,
total_sulfur_dioxide, sulphates, alcohol.

Log transformation did not improve: citric_acid, residual_sugar,
chlorides, free_sulfur_dioxide, and quality.

```{r}
par(mar = c(2, 2, 1, 1))
par(mfrow = c(3,2))
hist(log(fixed_acidity), cex.main =.8)
hist(log(volatile_acidity), cex.main =.8)
hist(log(total_sulfur_dioxide), cex.main =.8)
hist(log(sulphates), cex.main =.8)
hist(log(alcohol), cex.main =.8)
```

Using power transformation:

```{r}
par(mar = c(2, 2, 1, 1))
par(mfrow = c(2,2))
hist(sqrt(citric_acid), cex.main =.8)
hist((residual_sugar)^(-1/2), cex.main =.8)
hist((chlorides)^(-1/2), cex.main =.8)
hist((free_sulfur_dioxide)^(1/3), cex.main =.8)
```

```{r, echo = FALSE}
m1=lm(quality~log(fixed_acidity), data = wine_quality_red_data)
m2=lm(quality~density, data = wine_quality_red_data)
m3=lm(quality~ph, data = wine_quality_red_data)
m4=lm(quality~log(volatile_acidity), data = wine_quality_red_data)
m5=lm(quality~log(total_sulfur_dioxide), data = wine_quality_red_data)
m6=lm(quality~log(sulphates), data = wine_quality_red_data)
m7=lm(quality~log(alcohol), data = wine_quality_red_data)
m8=lm(quality~sqrt(citric_acid), data = wine_quality_red_data)
m9=lm(quality~I(residual_sugar^-(1/2)), data = wine_quality_red_data)
m10=lm(quality~I(chlorides^(-1/2)), data = wine_quality_red_data)
m11=lm(quality~I(free_sulfur_dioxide^(1/3)), data = wine_quality_red_data)
```

Thus, the final best transformation for the predictors are as follows:
log(fixed_acidity), density, ph, log(volatile_acidity),
log(total_sulfur_dioxide), log(sulphates), log(alcohol),
sqrt(citric_acid), (residual_sugar)\^(-1/2), (chlorides)\^(-1/2),
(free_sulfur_dioxide)\^(1/3)

\newpage

Below is the scatter plot matrix and q-q plots of the transformed
variables:

```{r, fig.width=20, fig.height=20}
n1 <- data.frame(log(fixed_acidity), log(volatile_acidity), sqrt(citric_acid), 
                 I(residual_sugar^-(1/2)), I(chlorides^-(1/2)), I(free_sulfur_dioxide^(1/3)), 
                 log(total_sulfur_dioxide), density, 
                 ph, log(sulphates), log(alcohol), quality)
plot(n1)
```

```{r, echo = FALSE}
residualsmodel1<-residuals(m1)
residualsmodel2<-residuals(m2)
residualsmodel3<-residuals(m3)
residualsmodel4<-residuals(m4)
residualsmodel5<-residuals(m5)
residualsmodel6<-residuals(m6)
residualsmodel7<-residuals(m7)
residualsmodel8<-residuals(m8)
residualsmodel9<-residuals(m9)
residualsmodel10<-residuals(m10)
residualsmodel11<-residuals(m11)
```

```{r, echo = FALSE}
par(mar = c(3, 3, 1.5, 1.5))
par(mfrow = c(3,4))
qqnorm(residualsmodel1)
qqnorm(residualsmodel2)
qqnorm(residualsmodel3)
qqnorm(residualsmodel4)
qqnorm(residualsmodel5)
qqnorm(residualsmodel6)
qqnorm(residualsmodel7)
qqnorm(residualsmodel8)
qqnorm(residualsmodel9)
qqnorm(residualsmodel10)
qqnorm(residualsmodel11)
```

Looking at the Scatterplot Matrix for the transformed predictors:
Because quality is numeric and not continuous, our graph when examining
the relationship between quality and the predictors do not appear like a
regular scatterplot. A better investigation would be using Generalized
Linear Models where the error terms would allow our dependent variable
to be continuous, though that is out of the scope of this class we
attempted the GLM later on in the report. In comparing the the
scatterplot matrix before and after the transformation, we will see that
the bunching of points on the lines appear to clump towards the center
and begin to sparse out at both ends of the line. Indicating that the
transformation indeed normalized the data.

Trying to use all these predictors will lead our model to suffer from
overfitting, so we must proceed with caution.

Looking at the scatterplot matrix and identifying collinearity, as well
as an intuitive understanding each predictor variable and their relation
to another on the physicochemical level, we decide to drop predictors:
fixed_acidity and total_sulfur_dioxide. Looking at our scatterplot
matrix of the transformed variable, we see high collinearity between
free_sulfur_dioxide vs. total_sulfur_dioxide, between density vs.
fixed_acidity, and between fixed acidity vs. ph. The multicollinearity
of fixed_acidity must be addressed first, and thus dropping the
predictor over density or ph is most logical. Furthermore, though each
predictors do not have strong linear relationship with the dependent
variable, we are interested in seeing when the dependent variables are
combined in a multilinear model, how much more could it help to explain
the association with the quality variable. Second, we see that there is
indepedence and no strong correlation between the predictors that are
not dropped. Third, we also checked for homoscedasticity and normality
with the histograms and q-q plots above, proving the constant variance
within our model. Thus, the 4 linear regression assumptions have been
checked to proceed with our model building.

#### Model building: Red wine dataset

We begin first with the full model after addressing the
multicollinearity issue.

```{r}
fullmodel<- lm(quality ~ density + ph + log(volatile_acidity) + log(sulphates) + log(alcohol) 
               + sqrt(citric_acid) + I(residual_sugar^-(1/2)) + I(chlorides^(-1/2)) 
               + I(free_sulfur_dioxide^(1/3)), 
               data = wine_quality_red_data)

```

Removal of predictors due to non-signifant p-values. Model 2: in model 2
we drop density and then conduct the anova(model2, fullmodel) test. With
the p value of 0.5877, which suggests density should be removed as we
fail to reject the NH. Then, we move to drop I(residual_sugar\^-(1/2))
due to a nonsignificant p-value. When conducting anova(model3, model2),
with a p-value of 0.4492, the test suggests residual sugar should be
removed as we fail to reject the NH. A summary of model 3 indicates to
drop I(free_sulfur_dioxide\^(1/3)), and as we conduct the anova(model4,
model3) with the removal of the predictor, we receive a p-value of
0.3082, which suggests free sulfur dioxide should be removed as we fail
to reject the NH. Finally, running the summary for model 4 shows that
all coefficients are significant and we can stop dropping predictors.

```{r}
model2<- lm(quality ~ ph + log(volatile_acidity) + log(sulphates) + log(alcohol) + 
              sqrt(citric_acid) + I(residual_sugar^-(1/2)) + I(chlorides^(-1/2)) 
            + I(free_sulfur_dioxide^(1/3)), data = wine_quality_red_data)
anova(model2, fullmodel)

model3<- lm(quality ~ ph + log(volatile_acidity) + log(sulphates) + log(alcohol) + 
              sqrt(citric_acid) + I(chlorides^(-1/2)) + I(free_sulfur_dioxide^(1/3)), 
            data = wine_quality_red_data)
anova(model3, model2)

model4<- lm(quality ~ ph + log(volatile_acidity) + log(sulphates) + log(alcohol) + 
              sqrt(citric_acid) + I(chlorides^(-1/2)), data = wine_quality_red_data)
anova(model4, model3)

summary(model4)
```

We arrive at Model4:lm(quality \~ ph + log(volatile_acidity) +
log(sulphates) + log(alcohol) + sqrt(citric_acid) + I(chlorides\^(-1/2))
after reducing our full model.

Added Variable Plot & Effect plot: From the model4, check the variable
'sqrt(citric_acid)' (significant at 0.01 level) From avp, we see the
residual of sqrt(citric_acid) regression on other variables shows
significance which proves it cover the left part of other variables to
quality and thus we should keep the sqrt(citric_acid). From effect plot, it is shows a linear plot.

```{r, echo = FALSE}
#add variable plot
model_avp1<- lm(quality ~ ph + log(volatile_acidity) + log(sulphates) + log(alcohol) 
                + I(chlorides^(-1/2)), data = wine_quality_red_data)
model_avp2<-lm(sqrt(citric_acid) ~ ph + log(volatile_acidity) + log(sulphates) 
               + log(alcohol) + I(chlorides^(-1/2)), data = wine_quality_red_data)
avp<-lm(residuals(model_avp1)~residuals(model_avp2))
summary(avp)
par(mfrow = c(2,2))
plot(avp)
#Effect plot
library(effects)
sqrt_citric_acid=sqrt(citric_acid)
model4_e<- lm(quality ~ ph + log(volatile_acidity) + log(sulphates) + log(alcohol) + sqrt_citric_acid + I(chlorides^(-1/2)), data = wine_quality_red_data)
plot(Effect('sqrt_citric_acid', model4_e))
```

\newpage

Thus, the final model for the ANOVA method of slowing dropping independent
variables is model 4. We now move on to check model4's residuals:
whether it is normally distributed to confirm homoscedasticity.

```{r}
ggplot(wine_quality_red_data,aes(model4$fitted.values,model4$residuals))+geom_point()
```

Specifically examining the residual vs. fitted values plot, we are able
to check for homoscedasticity. The above graph appears the way it does
because wine quality is not a continuous variable, whereby ratings can
only be whole numbers from a scale of 1 to 10. This enforce our fitted
values to be roughly whole numbers, hence the bands appear as such.

We are technically breaking the rule, as we are supposed to use a
continuous dependent variable in linear regression. However, because
wine quality is an increasing value, higher wine quality scores are
interpretable and meaningful to our analysis, and the range of the
scores are relatively large (1 to 10), and the distribution of the
scores varies reasonably across the range of available scores, we are
able to treat wine quality as a continuous variable. However, our
scatterplot will not show this, and what is observed are multiple linear
lines.

To interpret the above plot of residuals vs. fitted values, we look for
areas where the dots are bunched across the bands. We see bunching of
dots in specifically the middle bands, while lower or higher residual
bands do not show the bunching. This indicates that our assumption of
homoscedasticity, assumption of equal or similar variances in different
predictor variable groups, is satisfied.

Next, we use the working model and begin to examine outliers by looking
at Residuals and Leverage through Cook's distance, studentized
residuals, and hat_matrix.

First, we will investigate on a case by case simple linear regression
case for each predictor variable.

model4\<- lm(quality \~ ph + log(volatile_acidity) + log(sulphates) +
log(alcohol) + sqrt(citric_acid) + I(chlorides\^(-1/2)), data =
wine_quality_red_data)

ph: m3 log(volatile_acidity): m4 log(sulphates): m6 log(alcohol): m7
sqrt(citric_acid): m8 I(chlorides\^(-1/2)): m10

Determine Outliers, specifically on the chosen predictors. Regression
analysis to see the impact of those outliers. Leverage and residual
level.

Outliers Test: 'The function outlierTest from the car package gives the
most extreme observation based on the given model.'

```{r}
car::outlierTest(m3)
car::outlierTest(m4)
car::outlierTest(m6)
car::outlierTest(m7)
car::outlierTest(m8)
car::outlierTest(m10)
car::outlierTest(model4)
```

Running the outlierTest on the six predictors of our working model
individually, reveals the index point for investigation. m3: identifies
518, m4: identifies 391, m6: identifies 833, m7: identifies 900, m8:
identifies 460, m10: identifies 1470 Moreover, the overall regression
model identifies 833 as the most extreme observation based on the given
model.

We now move to examine these points closer, looking at
influenceIndexPlot to investigate the Cook's Distance, Studentized
residuals, and Hat_values. 
Cook's distance: The cook's distance providesthe estimate of the influence of a data point. By taking into account both leverage and residual of the observation, cook's distance is the summary of how much the regression model changes when the observation is removed.\
Studentized Residuals: The studentized residual is the residual divided by its standard deviation. Thus, they quantify how large the residuals are in standard deviation units, and allow for easier identification of outliers.\ 
Hat values: Hat values are the fitted values, the predictions made by the model for each observed values.\
In examining all three values of the Simple Linear Regression models individually, then the Multiple Linear Regression model, and taking into account all three values and the respective detected observations, a final list of outliers will be formed.\

Influence index Plots  to determine outliers:

```{r, fig.width=12, fig.height=5}

influenceIndexPlot(m3)
influenceIndexPlot(m4)
influenceIndexPlot(m6)
influenceIndexPlot(m7)
influenceIndexPlot(m8)
influenceIndexPlot(m10)
influenceIndexPlot(model4)

```

m3: high cook's distance for 152, 589\
m4: high hat_values for 950, high bonferroni p-values for 833, and high
studentized residual for 833\
m6: high cook's distance and hat_values for 152, high bonferroni
p-values for 152, 833\
m7: high hat_values and cook's distance for 653\
m8: no noticeable concern\
m10: high cook's distance for 282 and 1375/ High hat_values for 838

The decision is to remove the observations with index 152, 653, 833, and
838.

```{r}
to.rm <- c(152, 653, 833, 838)
newer_wine_quality_red_data <- wine_quality_red_data[-to.rm,]
rownames(newer_wine_quality_red_data) <- NULL

```

Examine how the removal of the 4 points will effect the each predictor
variable case by case. The only predictor variable whom R-squared value
did not improve was m10: regressing quality on I(chlorides\^(-1/2)).

```{r, echo = FALSE}
m3newer<- lm(quality~ph, data = newer_wine_quality_red_data)

m4newer=lm(quality~log(volatile_acidity), data = newer_wine_quality_red_data)

m6newer=lm(quality~log(sulphates), data = newer_wine_quality_red_data)

m7newer=lm(quality~log(alcohol), data = newer_wine_quality_red_data)


m8newer=lm(quality~sqrt(citric_acid), data = newer_wine_quality_red_data)


m10newer=lm(quality~I(chlorides^(-1/2)), data = newer_wine_quality_red_data)

```

New m3: R-squared value of new model: 0.003963 vs old model's: 0.003333\
m3 is lm(quality\~ph)\

New m4: R-squared value of new model: 0.1546 vs old model's: 0.1531\
m4 is lm(quality\~log(volatile_acidity)\

New m6: R-squared value of new model: 0.1063 vs old model's: 0.09526\
m6 is lm(quality\~log(sulphates)\

New m7: R-squared value of new model: 0.02308 vs old model's: 0.2275\
m7 is lm(quality\~ph)\

New m8: R-squared value of new model: 0.04556 vs old model's: 0.04272\
m8 is lm(quality\~sqrt(citric_acid)\

New m10: R-squared value of new model: 0.03159 vs old model's: 0.03593\
m10 is lm(quality\~I(chlorides\^(-1/2))\

Now examine the full model with those outliers removed:

```{r}
model4newer<- lm(quality ~ ph + log(volatile_acidity) + log(sulphates) + log(alcohol) 
                 + sqrt(citric_acid) + I(chlorides^(-1/2)), 
                 data = newer_wine_quality_red_data)
summary(model4newer)

```

Model4newer R-squared value is 0.3625, whereas Model4 R-squared value is
0.3541. It seems that the new model performs more reliable. However,
outliers correction can only do so much, as a 0.3625 R-squared value is
till quite low.

Perhaps we can consider dropping the I(chlorides\^(-1/2)) predictor in
general. We will conduct an F-test of the full model and the lesser
model.

```{r}
model5newer<- lm(quality ~ ph + log(volatile_acidity) + log(sulphates) + log(alcohol) 
                 + sqrt(citric_acid), data = newer_wine_quality_red_data)
anova(model5newer, model4newer)
```

The anova test suggests that we should not remove I(chlorides\^(-1/2)),
as we reject the Null Hypothesis. It would be better to leave the
predictor in our model, and accept the removal of the 4 outliers, though
decreased the R-Squared value within the simple linear regression of
quality onto I(chlorides\^(-1/2)), a view on the multiple linear
regression level suggests the removal of the outliers contributes better
to the overall model. We will leave I(chlorides\^(-1/2)) within the full
model and move accept fully the elimination of the four outliers.

Final Model #1 for Red Wine dataset:

```{r}
model4newer$coefficients
```

The ANOVA method is not the sole method to build the best models. Below,
we will consider multiple different approaches.

Another Model: Interaction terms\
We did not choose to implement interaction terms in our method because
we believed the problems of multicollinearity between our independent
variables would be further complicated when creating interaction terms.
Since we decide to drop the two predictors with high issues of
multicollinearity, we do not feel it necessary to approach and
complicate the model further with interaction terms, they are simply not
of interest for our model. And because we do not have any categorical
predictors, we do not see the implementation to produce an effective
model.

Another Model: In terms of using a method such as Weighted Least
Squares, we believe the usage of Ordinary Least Squares is sufficient in
this project due to the constant variance achieved through our
transformation.

Another model: Forward selection with transformed variables for AIC
(Akaike Information Criterion)

```{r, echo = FALSE}
TRfixed_acidity = log(wine_quality_red_data$fixed_acidity)
TRvolatile_acidity=log(wine_quality_red_data$volatile_acidity)
TRcitric_acid = sqrt(wine_quality_red_data$citric_acid)
TRresidual_sugar=(wine_quality_red_data$residual_sugar)^(-1/2)
TRchlorides = (wine_quality_red_data$chlorides)^(-1/2)
TRfree_sulfur_dioxide = (wine_quality_red_data$free_sulfur_dioxide)^(1/3)
TRtotal_sulfur_dioxide = log(wine_quality_red_data$total_sulfur_dioxide)
Tdensity=wine_quality_red_data$density
Tph=wine_quality_red_data$ph
TRsulphates = log(wine_quality_red_data$sulphates)
TRalcohol = log(wine_quality_red_data$alcohol)
Tquality = wine_quality_red_data$quality


TRwine_quality_red_data =  data.frame(TRfixed_acidity, TRvolatile_acidity, TRcitric_acid, TRresidual_sugar, TRchlorides, TRfree_sulfur_dioxide, TRtotal_sulfur_dioxide, Tdensity, Tph, TRsulphates, TRalcohol, Tquality)

#define intercept-only model
intercept_only <- lm(Tquality ~ 1, data=TRwine_quality_red_data)

#define model with all predictors
all <- lm(Tquality ~ ., data=TRwine_quality_red_data)
forward <- step(intercept_only, direction='forward', scope=formula(all), trace=1)
```

Our AIC forward selection gives the lowest AIC value for the predictors
and give the final model as:

```{r}
forward$anova
forward$coefficients
```


Another model: Backward Elimination AIC with the transformed variables

```{r, echo = FALSE}

backward <- step(all, direction='backward', scope=formula(all), trace=1)
summary(backward)
```


Our AIC backward selection gives the lowest AIC value for the predictors
and give the final model as:

```{r}
backward$anova
backward$coefficients
```


Another model: Stepwise with the transformed variables, allowing us to
reduce the full model down to the following.

```{r, echo = FALSE}
library(MASS)
model_transformed <- lm(quality ~ log(fixed_acidity) + log(volatile_acidity) + sqrt(citric_acid) + I(residual_sugar^(-0.5)) + I(chlorides^(-0.5)) + I(free_sulfur_dioxide^(-0.5)) + log(total_sulfur_dioxide) + density + ph + 
            log(sulphates) + log(alcohol) , data = wine_quality_red_data)
summary(model_transformed)
step.modeltr <- stepAIC(model_transformed, direction = "both", 
                      trace = FALSE)
```

```{r}
summary(step.modeltr)
```


Another model: Forward selection with transformed variables for BIC
(Bayesian Information Criterion)

```{r, echo = FALSE}
sample_size <- nrow(wine_quality_red_data)
BIC_modelfor <- stepAIC(fullmodel, direction = "forward",
         criterion = c("BIC"),
         trace = 0,
         k = log(sample_size))
BIC(BIC_modelfor)
summary(BIC_modelfor)
```


Our BIC forward selection gives:

```{r}
BIC_modelfor$anova
BIC_modelfor$coefficients
```


Another model: Backward Elimination BIC with the transformed variables

```{r, echo = FALSE}
BIC_modelback <- stepAIC(fullmodel, direction = "backward",
         criterion = c("BIC"),
         trace = 0,
         k = log(sample_size))
BIC(BIC_modelback)
summary(BIC_modelback)
```

```{r}
BIC_modelback$anova
BIC_modelback$coefficients
```


Transformed BIC Model w/ Stepwise

```{r, echo = FALSE}
library(MASS)
BIC(fullmodel)
sample_size <- nrow(wine_quality_red_data)
BIC_model <- stepAIC(fullmodel, direction = "both",
         criterion = c("BIC"),
         trace = 0,
         k = log(sample_size))
BIC(BIC_model)
summary(BIC_model)
```

```{r}
BIC_model$anova
BIC_model$coefficients
```

\newpage
#### The seven Final models:

```{r}

coefficients(model4newer)
coefficients(forward)
coefficients(backward)
coefficients(step.modeltr)
coefficients(BIC_modelback)
coefficients(BIC_modelfor)
coefficients(BIC_model)

```

#### Select Best Model:

We use K-Fold Cross Validation to decided the best model and specify the
cross-validation method (repeat 3 times).

```{r, echo = FALSE}
library(caret)

ctrl <- trainControl(method = "repeatedcv",
                            number = 10, repeats = 3)

model_cv1 <- train(quality ~ ph + log(volatile_acidity) + log(sulphates) + 
    log(alcohol) + sqrt(citric_acid) + I(chlorides^(-1/2)), data = newer_wine_quality_red_data, method = "lm", trControl = ctrl)

model_cv2 <- train(Tquality ~ TRalcohol + TRvolatile_acidity + TRsulphates + 
    TRchlorides + Tph + TRtotal_sulfur_dioxide + TRfree_sulfur_dioxide, 
    data = TRwine_quality_red_data, method = "lm", trControl = ctrl)

model_cv3 <- train(Tquality ~ TRfixed_acidity + TRvolatile_acidity + 
    TRcitric_acid + TRresidual_sugar + TRchlorides + TRfree_sulfur_dioxide + 
    TRtotal_sulfur_dioxide + Tdensity + Tph + TRsulphates + TRalcohol, 
    data = TRwine_quality_red_data, method = "lm", trControl = ctrl)

model_cv4 <- train(quality ~ log(volatile_acidity) + I(chlorides^(-0.5)) + 
    I(free_sulfur_dioxide^(-0.5)) + log(total_sulfur_dioxide) + 
    ph + log(sulphates) + log(alcohol), data = wine_quality_red_data, method = "lm", trControl = ctrl)
```

Viewing the summary of k-fold CV, the best fit model should have small
RMSE&MAE and large R-square.

```{r}
par(mfrow = c(2,2))
print(model_cv1)
print(model_cv2)
print(model_cv3)
print(model_cv4)
```

Therefore, our best model is

```{r}
summary(model4newer)
```

Furthermore, Looking at the R-squared values for the above models, the
range of the R-squared values are between 0.3522 and 0.3625. With our
two best models, providing the two highest R-squared values being:
model4newer, from ANOVA, and step.modeltr, from AIC analysis.

#### Alternative methods: General Linear Regression

In considering our response variable quality is in integer scaled and
skewed residual plot of variables. Our group advised that a try of using
generalized linear model can help. Instead of requiring strictly normal
distribution of residual, generalized linear model allows a variety of
other distributions from the exponential family of residual. Here, we
try logistic regression as a family of generalized linear model.

Logistic regression fits into a sigmoid function might be good.
Different from the assumptions of linear regression, logistic does not
require to hold linearity, normality, and homoscedasticity. However,
dependent variable needs to be binary and ordinal.

Here we advise two ways to run the logistic regression. In the first
method, quality is converted into binomial variable. Inspired by
original scatter plots of quality and predictors, we notice that quality
valued at 6 is like a watershed of inferior and superior. Moreover,
statistically proved by the summary of quality column, we can see 6 is
the median. Thus, in the first method, we convert quality into inferior
and superior group. It helps us to answer our guiding question: what
physicochemical properties influence the quality score of wine?

For the second method, categorical variable type of wine in white or red
becomes response variable in logistic regression. We believe it will
benefit us to answer our guiding question that whether we can
differentiate red and white wine by proportion of chemicals it contains?

First method below:

```{r}
redw <- read.csv(file="wine_quality_red.csv")
whitew <- read.csv(file="wine_quality_white.csv")

summary(redw$quality)
redw$new = ifelse(redw$quality>6,1,0)
```

At first, we fit the full model, where we have AIC 894.88.

```{r}
reg <- glm(new~fixed_acidity+volatile_acidity+citric_acid+residual_sugar+chlorides+free_sulfur_dioxide+total_sulfur_dioxide+density+ph+sulphates+alcohol,redw,family = binomial)
summary(reg)
```

Then, we drop the highly-correlated variables and non-significant
variables. We reach the model below.

```{r, echo = FALSE}
reg2 <- glm(formula = new ~ fixed_acidity + volatile_acidity + residual_sugar + fixed_acidity + density + sulphates + alcohol, data = redw,family = binomial)
```

```{r}
summary(reg2)
1-pchisq(1269.92-872.08,8)
```

Interpretation toward result of logistic regression: the null deviance
tells us how well the response variable can be predicted by a model with
only intercept. The residual deviance shows how well the response can be
predicted when predictors are included.

The difference between the two deviance is dropping deviance with the
chi-squared distribution. Our null: there is no difference in a
deviation between null deviance and residual deviance. Based on the
p-value of 0, we fail to reject the null that there is a difference in
deviance.

Second method below:

```{r}
redw <- read.csv(file="wine_quality_red.csv")
redw$redwine = 1
whitew$redwine = 0
wine <- rbind(redw,whitew)
reg3 <- glm(redwine~fixed_acidity+volatile_acidity+citric_acid+residual_sugar+chlorides+free_sulfur_dioxide+total_sulfur_dioxide+density+ph+sulphates+alcohol+quality,wine,family = binomial)
summary(reg3) #450

reg4 <- glm(redwine~volatile_acidity+citric_acid+residual_sugar+chlorides+free_sulfur_dioxide+total_sulfur_dioxide+density+sulphates+alcohol,wine,family = binomial)
summary(reg4) #451
1-pchisq(7250.98-431.73,9)
```

For the second method, we combine two datasets and use categorical
variable white or red wine as response variable for logistic regression.

As result, we fail to reject the null that there is a difference in
deviance.

Conclusion for trying generalized linear model(GLM): Two logistic
regressions that we tried are not meaningful from the p-value of the
chi-squared test. The failure might cause by three possibilities. First
is the ignorance of GLM assumptions. Second, the standard that we used
for dropping variables is collinearity and significance. There might be
better stats or tests to use as standards for dropping. Third, utilizing
the p-value from the chi-squared test to decide may not be suitable or
best for deciding the significance of the whole model. We expected to
learn more about GLM in the future.

\newpage
#### Model building: White wine dataset

We now proceed to analyze our White wine samples in the same manner as
we have analyzed the Red wine samples. Instead of implementing all 7
models, we will implement the analysis of the two best red wine models
on our white wine dataset. These two best models were through ANOVA and
AIC analysis.

```{r, echo = FALSE}
wine_quality_white_data <- read.csv(file="wine_quality_white.csv",as.is=T, header=T)
w <- data.frame(fixed_acidity = wine_quality_white_data$fixed_acidity, volatile_acidity = wine_quality_white_data$volatile_acidity, citric_acid = wine_quality_white_data$citric_acid, residual_sugar = wine_quality_white_data$residual_sugar, chlorides = wine_quality_white_data$chlorides, free_sulfur_dioxide = wine_quality_white_data$free_sulfur_dioxide, total_sulfur_dioxide = wine_quality_white_data$total_sulfur_dioxide, density = wine_quality_white_data$density, ph = wine_quality_white_data$ph, sulphates = wine_quality_white_data$sulphates, alcohol = wine_quality_white_data$alcohol, quality = wine_quality_white_data$quality)
```

```{r, fig.width=15, fig.height=15}
plot(w) 

```

Below shows the histogram of our transformed predictors:

```{r, echo = FALSE}
par(mar = c(2, 2, .5, .5))
par(mfrow = c(2,2))
hist(log(wine_quality_white_data$fixed_acidity), cex.main =.7)
hist(log(wine_quality_white_data$volatile_acidity), cex.main =.7)
hist(sqrt(wine_quality_white_data$total_sulfur_dioxide), cex.main =.7) 
hist(log(wine_quality_white_data$ph), cex.main =.7)
```

```{r, echo = FALSE}
par(mar = c(3, 3, 0.5, 0.5))
par(mfrow = c(3,2))
hist(log(wine_quality_white_data$sulphates), cex.main =.8)
hist((wine_quality_white_data$alcohol)^(-1/2), cex.main =.8) 
hist(sqrt(wine_quality_white_data$citric_acid), cex.main =.8)
hist(sqrt(wine_quality_white_data$residual_sugar), cex.main =.8)
hist((wine_quality_white_data$chlorides)^(-1/3), cex.main =.8)
hist((wine_quality_white_data$free_sulfur_dioxide)^(1/2), cex.main =.8)
```

Important to note is the omission of the histogram of the variable
density from the plots above. We decided to drop the predictor density
because it is highly correlated with residual_sugar, and we have tried
all power transformation on density, yet no transformation normalized
it. Thus, we arrive at the full model:

```{r}
fullmodelW<- lm(wine_quality_white_data$quality ~ log(wine_quality_white_data$ph) 
                + log(wine_quality_white_data$volatile_acidity) 
                + log(wine_quality_white_data$sulphates) 
                + I(wine_quality_white_data$alcohol^(-1/2)) 
                + sqrt(wine_quality_white_data$citric_acid) 
                + sqrt(wine_quality_white_data$residual_sugar) 
                + I(wine_quality_white_data$chlorides^(-1/3)) 
                + I(wine_quality_white_data$free_sulfur_dioxide^(1/2)))
coefficients(fullmodelW)
```

\newpage
Next, begin reducing the model through the removal of predictors due to
non-signifant p-values. The above model shows that the coefficient of
predictor citric acid is nonsignificant to our model.

```{r, echo = FALSE}
model2W<- lm(wine_quality_white_data$quality ~ log(wine_quality_white_data$ph) + log(wine_quality_white_data$volatile_acidity) + log(wine_quality_white_data$sulphates) 
             + I(wine_quality_white_data$alcohol^(-1/2)) + sqrt(wine_quality_white_data$residual_sugar) + I(wine_quality_white_data$chlorides^(-1/3)) 
             + I(wine_quality_white_data$free_sulfur_dioxide^(1/2)))
anova(model2W, fullmodelW)
```

Performing the ANOVA test suggests to drop the predictor citric acid.
Given the p-value of 0.9383, we fail to reject the NH.

Thus, all predictors, after the removal of the predictors density and
citric acid, are now significant, allowing the analysis to reach the
final model with intercept and coefficients:

```{r}
coefficients(model2W)
```

\newpage
Outliers check: Similar to the methods for outliers checking in the red
wine data set, the same is implemented for white wine.

Below are the simple linear regression models checked that will be
checked individually for outliers.

```{r}
w1=lm(wine_quality_white_data$quality~log(wine_quality_white_data$ph))
w2=lm(wine_quality_white_data$quality~
        log(wine_quality_white_data$volatile_acidity))
w3=lm(wine_quality_white_data$quality~log(wine_quality_white_data$sulphates))
w4=lm(wine_quality_white_data$quality~I(wine_quality_white_data$alcohol^(-1/2)))
w5=lm(wine_quality_white_data$quality~sqrt(wine_quality_white_data$residual_sugar))
w6=lm(wine_quality_white_data$quality~I(wine_quality_white_data$chlorides^(-1/3)))
w7=lm(wine_quality_white_data$quality~
I(wine_quality_white_data$free_sulfur_dioxide^(1/2)))
```

The influenceIndexPlot to examine Cook's Distance, Studentized
Residuals, and Hat_Values

```{r, fig.width=9, fig.height=5}
influenceIndexPlot(w1)
influenceIndexPlot(w2)
influenceIndexPlot(w3)
influenceIndexPlot(w4)
influenceIndexPlot(w5)
influenceIndexPlot(w6)
influenceIndexPlot(w7)
```

For w1: high cook's distance at 1689 and 251\
For w2: high cook's distance at 28, high hat_values at 4040\
for w3: no outliers detected\
for w4: high cook's distance at 3902 and 741\
for w5: high hat_values at 2782\
for w6: high cook's distance at 3774 and 3902\
for w7: high cook's distance and hat_values at 4746\

```{r, fig.width=9, fig.height=5}
influenceIndexPlot(model2W)
```

And directly above is the influenceIndexPlot for the multiple linear
regression model, which shows high cook's distance and hat_values at
4746, and high hat_values at 2782.

The decision is to remove the observations with index 28, 251, 741,
1689, 2782, 3774, 3902, 4040, 4746.

```{r}
to.remove <- c(1689, 251, 28, 4040, 3902, 741, 2782, 4746, 3774)
newer_wine_quality_white_data <- wine_quality_white_data[-to.remove,]
rownames(newer_wine_quality_white_data) <- NULL
```

```{r, echo = FALSE}
model2Wnewer<- lm(newer_wine_quality_white_data$quality ~ log(newer_wine_quality_white_data$ph) + log(newer_wine_quality_white_data$volatile_acidity) + log(newer_wine_quality_white_data$sulphates) + I(newer_wine_quality_white_data$alcohol^(-1/2)) + sqrt(newer_wine_quality_white_data$residual_sugar) + I(newer_wine_quality_white_data$chlorides^(-1/3)) + I(newer_wine_quality_white_data$free_sulfur_dioxide^(1/2)))

summary(model2Wnewer)

```

After dropping the outliers, the R-squared values of the regression
model improved from 0.2742 to 0.2795. This demonstrates that the removal
of these outliers of potentially high residual and high leverage has
improved the model. Next, we move to our second best model analysis
method: AIC.

AIC Stepwise variable selection:

```{r, echo = FALSE}
library(MASS)
winequality_white = wine_quality_white_data
colnames(winequality_white)=c("fixed.acidity", "volatile.acidity", "citric.acid",        "residual.sugar", "chlorides", "free.sulfur.dioxide","total.sulfur.dioxide", "density", "pH", "sulphates", "alcohol", "quality")
model_transformed <- lm(quality ~ log(fixed.acidity) + log(volatile.acidity) + sqrt(citric.acid) + sqrt(residual.sugar) + I(chlorides^(-1/3)) + sqrt(free.sulfur.dioxide) + sqrt(total.sulfur.dioxide) + density + log(pH) + 
            log(sulphates) + I(alcohol^(-0.5)) , data = winequality_white)
summary(model_transformed)

step.modelW <- stepAIC(model_transformed, direction = "both", 
                      trace = FALSE)
summary(step.modelW)
cat("The AIC value of the final model is ", extractAIC(step.modelW)[2], "with", extractAIC(step.modelW)[1]-1,"regressors")
```

```{r}
step.modelW$anova
step.modelW$coefficients 
```

Forward variable selection

```{r, echo = FALSE}
TRfixed_acidity = log(winequality_white$fixed.acidity)
TRvolatile_acidity=log(winequality_white$volatile.acidity)
TRcitric_acid = sqrt(winequality_white$citric.acid)
TRresidual_sugar=(winequality_white$residual.sugar)^(1/2)
TRchlorides = (winequality_white$chlorides)^(-1/3)
TRfree_sulfur_dioxide = (winequality_white$free.sulfur.dioxide)^(1/2)
TRtotal_sulfur_dioxide = sqrt(winequality_white$total.sulfur.dioxide)
Tdensity=winequality_white$density
Tph=log(winequality_white$pH)
TRsulphates = log(winequality_white$sulphates)
TRalcohol = (winequality_white$alcohol)^(-1/2)
Tquality = winequality_white$quality


TRwinequality_white =  data.frame(TRfixed_acidity, TRvolatile_acidity, TRcitric_acid, TRresidual_sugar, TRchlorides, TRfree_sulfur_dioxide, TRtotal_sulfur_dioxide, Tdensity, Tph, TRsulphates, TRalcohol, Tquality)

#define intercept-only model
intercept_only <- lm(Tquality ~ 1, data=TRwinequality_white)

#define model with all predictors
allW <- lm(Tquality ~ ., data=TRwinequality_white)
forwardW <- step(intercept_only, direction='forward', scope=formula(allW), trace=0)
summary(forwardW)

step(intercept_only, direction='forward', scope=formula(all), trace=1)
cat("The AIC value of the final model is ", extractAIC(forwardW)[2], "with", extractAIC(forwardW)[1]-1,"regressors")
```

```{r}
forwardW$anova
forwardW$coefficients

```

Backward Elimination:

```{r, echo = FALSE}
backwardW <- step(all, direction='backward', scope=formula(allW), trace=1)
summary(backwardW)
cat("The AIC value of the final model is ", extractAIC(backwardW)[2], "with", extractAIC(backwardW)[1]-1,"regressors")
```

```{r}
backwardW$anova
backwardW$coefficients
```
For the data related to white, the 3 variable selection methods shown above that use the AIC criterion give models with all the 11 regressors, i.e. without eliminating any of the regressors. However, the R-squared values produced is only 0.2869 and has an AIC value of -2822.955.



We have come to two best models, each for the two data sets. Below, we
will discuss and interpret the models on a deeper level and work our way
to answering our guiding question.

#### Discussion toward red wine final model

```{r, echo = FALSE}
summary(model4newer)
```

Based on our final choice of model, ph, log(volatile_acidity),
log(sulphates), log(alcohol), square(citric_acid), and negative square
root(chlorides). We believe these six variables explain our red wine
quality best in our many tries. In our final model for red wine, looking
at residuals, two pairs of max and min, 75%(3Q) and 25%(1Q) lines are
almost symmetric. It shows that our data points spread fairly well
validating with residual plot before. Viewing the slope sign of
variables, we notice that intercept, ph, log(volatile_acidity), and
square(citric_acid) have negative correlation with quality score. Higher
ph might bring a bitter taste. Statistically, keeping all other
variables the same, one increasing unit in ph will decrease the quality
grades by 0.61499. Higher volatile acidity brings unpleasant vinegar
taste. Keeping other variables same, with one percentage increase in
volatile acidity, the grades of wine quality will lower by 0.53754. Both
ph and volatile acidity fit with logical sense. Surprisingly, citric
acid as adding freshness and flavor to the wine is supposed to be
positive. The actual negative sign shows the fact that the large amount
of citric acid might bring an unpleasant taste to the wine instead of
making it better.

Similarly, one-percent increase in sulphates and alcohol leads to
increases in quality rank by 0.77403 and 3.27477 respectively. One more
unit square root of citric acid will decrease the ranking of wine
quality by 0.21635 units. One more unit inverse of square root of
chlorides will increase the ranking of wine quality by 0.11452. Among
three components that make positive impacts on the red wine quality,
alcohol brings the strongest positive effects. People might generally
favor more alcohol containing wine. More SO2 in the wine also has a
positive influence. Chlorides represent the amount of salt in the wine.
Even it has a positive sign, reminding the reciprocal, smaller chlorides
and less salty taste of wine will have a higher quality score.

Furthermore, besides intercept, we get almost all included explanatory
variables significantly. Note that the non-significant intercept does
not pose an issue to the validity of our model. The non-significant
intercept is quite meaningless given that not all our predictors will
take on the value of 0, thus whether we have sufficient statistical
evidence that the intercept is different from zero is not relevant
information. Though, it might infer that perhaps we need more relevant
data to better predict the wine quality score, besides the given
chemical compositions. From adjusted R-squared, we know about 36.25% of
the variability in quality score between different wines can be
explained by the linear association that we display in this final model.
Here, we use adjusted one instead of multiple R-squared to take account
of overfitting. Our fairly small p-value of F-stats show the stronger
evidence against null. In other words, the smaller probability of
getting an F-statistic greater than the value actually observed.

#### Discussion toward white wine final model

The best models forn white wine were the 3 variable selection methods shown above that use the AIC criterion give models with all the 11 regressors, i.e. without eliminating any of the regressors. However, the R-squared values produced is only 0.2869 and has an AIC value of -2822.955. Whereas the ANOVA method for white wine only 0.2795. 

#### Conlusion

Wine quality scores are related to chemical compositions. A lower level
of volatile acidity, chlorides,citric acid, pH and higher level of
sulphates and alcohol could generally result in better wine tasting.

Since both models for red and white wine yielded very small p-values
they cannot support the hypothesis that the physicochemical properties
of wine have an influence on the resulting quality. However, we can
still gain some insight about the differences between our red and white
wine models by comparing their summaries. Red wine had a noticeably
higher R-squared value of 0.363 compared to whites 0.28. The two models
share 4 variables in common of which 2 utilized the same transformation,
volatile acidity & sulphates. The estimated coefficient for volatile
acidity is very similar between models with red having -0.5375 and white
-0.56284.

#### Motivation and Further Improvements

Beyond the general understanding that alcohol is in wine, two wine
quality datasets provide us with a professional view into chemical
composition of wine. Moreover, quality of wine can possibly related with
serious chemistry. All those new aspects attract us to choose wine
quality datasets. While we browsed the dataset at first time, we notice
the explantory variables of homogeneous chemical compositions are
usually contrasted sharply with human tasting score. Because chemistry
can represent rational, human tasting is more about emotional. The huge
difference and possible difficulties that we might encounter in the
analysis further motivate us to use this dataset and face the
challenges.

As we perceived the challenges when we picked up the dataset, the whole
project brings surprising growth for every member in our group. We have
a consensus that group project explore our world to more than simple or
multivariate linear model. The class-based knowledge is more familiar to
us by discussing and questioning in group discussion. The difficulty of
handling real data give our a warning to choose and constructing model
more carefully.

For our analysis, there are several possible further improvements.
First, there are more than 4898 samples of white wine, whereas there are
only 1599 samples of red wine available. If there were more resources,
the group would wish to collect more samples of red wine, matching the
number of white wine samples, in order to carry out a more fair analysis
in comparing the two types of wine as well as more accuracy for the red
wine analysis.

Second, the quality data is median of 3 tasters. Quality score might be
biased and implicitly influence our final model.

Third, as the wine samples are from the north of Portugal, with further
resources, we would hope to add other variables such as the specific
regions of the north of Portugal the samples belong to, the price of the
wine bottle of the samples, and the years the wine samples were made, as
these are very important variables which affect how one would attribute
to the quality of the wine. An even more crucial variable would be the
type of grapes that were used for the wine samples, as this would
greatly influence the physicochemical elements and color of the
resulting wine. Perhaps the white wine samples were all made from a
specific type of yellow grape, while all the red wine samples were from
a specific type of purple grape; we would like to collect this
information to understand the biases of our samples.

Fourth, there are more analysis types that we can use to approach, such
as time series analysis. If we can historical data to predict the wine
quality. We can see systemic patterns in the quality of particular wine
quality of the region. This will allow us to investigate seasonal
changes in accordance with wine harvesting and wine quality.

\
\
\
\
